Aspect-oriented RTL HW design using SystemC

### 2014-01-26 Papers_SVN/papers/submitted/micpro-aspects_systemc-tiago/paper.tex
Versionando o meu mestrado.

%\documentclass[10pt,journal,final,letterpaper]{IEEEtran}
%\documentclass[10pt,peerreview,final,letterpaper]{IEEEtran}
\documentclass[prodmode,acmtodaes]{acmsmall}

\usepackage[latin1]{inputenc}	% for Latin languages
\usepackage[T1]{fontenc}	% for ISO and UTF characters
\usepackage[english]{babel}	% for multilingual support
\usepackage{graphicx}
%\usepackage{multirow}
%\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{color}
\definecolor{Red}{rgb}{.9,0,0}

@@

@@

\markboth{T. R. Mück and A. A. Fröhlich}{AOP-based RTL HW design using SystemC}

\title{AOP-based RTL HW design using SystemC}

%\author{Tiago Rogério Mück and Antônio Augusto Fröhlich\\
%\IEEEauthorblockA{Software/Hardware Integration Lab\\
%Federal University of Santa Catarina\\
%Florianópolis, Brazil\\ 
%Email: \{tiago,guto\}@lisha.ufsc.br}
%}

\author{TIAGO ROGÉRIO MÜCK and ANTÔNIO AUGUSTO FRÖHLICH
\affil{Federal University of Santa Catarina}}

@@

%\begin{IEEEkeywords}
%Aspect-oriented Programming, Digital Hardware Design, Hardware Description Languages, SystemC
%\end{IEEEkeywords}

\category{B.6.3}{LOGIC DESIGN}{Hardware description languages}

\terms{Design, Languages}

\keywords{Aspect-oriented Programming, SystemC}

@@

\begin{bottomstuff}
This work is partially supported by the \emph{Coordination for Improvement of Higher Level
Personnel}~(CAPES), under grants RH-TVD 006/2008 and 240/2008.

Author's addresses: T. R. Mück and A. A. Fröhlich, Software/Hardware Integration Lab, Federal
University of Santa Catarina, Florianópolis, Brazil, {tiago,guto}@lisha.ufsc.br.
\end{bottomstuff}

\maketitle

\section{Introduction}
\label{INTRO}
%The IEEE TII refenrences and some sentences were added tageting a submission to TII
The complexity of digital hardware design is increasing as the advances of the semiconductor
industry allow for the use of sophisticated computational resources in a wider range of
applications.
Low-power and small \textit{Field-programmable Gate Arrays}~(FPGAs) now reaching the market are one
of the products of these advances, enabling a more efficient and flexible implementation of
several industrial control systems that were previously based on microcontrollers
and even analog circuits~\cite{Monmasson:2007}.%also Monmasson:2011

This new deployment scenario is leading to a growing interest in high-level methodologies for
digital hardware design. Solutions and methodologies that have been successfully deployed in the
scope of large-scale software systems, such as \textit{object-oriented programming}~(OOP), are being
introduced to hardware design as well. An example of a \textit{hardware description language}~(HDL)
which supports OOP is SystemC, a C++-based HDL~\cite{Panda:2001}.

HDLs are used to create descriptions of electronic circuits. These descriptions can be used
either for design verification or for hardware synthesis. Differently from most software programming
languages, HDLs are intrinsically parallel and provide explicit means to describe timing.
VHDL~\cite{VHDL:2000} and Verilog~\cite{Verilog:2001} are the most widely used HDLs for 
\textit{register transfer level}~(RTL) design. In RTL, circuits are described in terms of the
operations between storage elements which are synchronized using clock signals. In this level, a
hardware design may be composed by several \emph{modules}, whose instances are interconnected to
build the system. However, in contrast to software programming languages, module communication
occurs through timed protocols and signals defined by the module input/output interface, instead of
function calls.

The use of OOP-capable HDLs (e.g. SystemC) enables an increased level of flexibility and reusability
in hardware design; however, analogous to software, in hardware some system-wide cross-cutting
concerns still cannot be elegantly encapsulated. For example, in complex circuits, interconnection
of several entities is realized by introducing buses. A bus physically interacts with other
components (e.g. CPU, DMA ...), but it is difficult to use a module or an OOP class to encapsulate
the bus because its interface and arbitration method have to be implemented in every attached
component~\cite{Engel:2008}. Other examples of crosscutting concerns in hardware design can be
found in parts of a system related to its overall functionality or the implementation of
non-functional properties (e.g. fault-tolerance and low-power mechanisms, hardware debugging through
scan chains, clock handling, ...)~\cite{Endoh:2011}. Even by using OOP in hardware, this scattered
code is hard to maintain and bugs may be easily introduced. This motivates the introduction of
\emph{aspect-oriented programming}~(AOP)~\cite{Kiczales:1997} techniques for hardware design. AOP
is an elaboration over OOP to deal with crosscutting concerns. AOP proposes the encapsulation of
these concerns in special classes called \emph{aspects}. An aspect can alter the behavior of the
base code by applying \emph{advices} (small pieces of code defining additional behavior) at specific
points of a program called \emph{join points}, which are specified in units called \emph{pointcuts}.

@@

In order to contribute to this scenario, in this paper we describe a digital hardware design method
which leverages on SystemC features in order to enable the implementation of hardware components
using OOP and AOP concepts. We propose the use of a domain engineering strategy which yields
components in which its dependencies from the execution scenario are encapsulated as \emph{aspects}
and \emph{configurable features}. These artifacts can be implemented using only standard
C++~\cite{Czarnecki:2000}, hence eliminating the need of extra tools and compilers. Additionally,
such features are within the SystemC synthesizable subset~\cite{systemc_subset}, thus yielding
\emph{synthesizable components}. Our method is finally illustrated by the design and implementation
of case studies extracted from an industrial \textit{Private Automatic Branch Exchange} (PABX)
application.

@@

There are also several works which proposed the use of AOP concepts mostly for 
hardware verification. \emph{Kallel et al.}~\cite{Kallel:2010} proposed the use of SystemC and
AspectC++~\cite{Spinczyk:2002} to implement assertion checkers. The authors focused on the
verification of
\textit{transaction-level models}~(TLM)~\cite{Cai:2003} in which transaction state updates are used
as
pointcuts. They provide a framework in which the user's verification classes extend base aspect
classes that implement the pointcuts and the verification primitives. In \emph{Vachharajani et
al.}~\cite{Vachharajani:2004} the authors developed the \textit{Liberty Structural
Specification Language}~(LSS). In LSS each module can declare instances which emit certain
events at runtime. These events behave like pointcuts of AOP. Each time a certain state is reached
or a value is computed, the instance will emit the corresponding event and user-defined aspects will
perform statistics calculation and reporting. \emph{Liu et al.}~\cite{Liu:2010} also proposed
AOP-based instrumentation, but focusing on high-level power estimation. They have developed a
methodology based on SystemC in which AspectC++ is used to define special power-aware aspects. 
These aspects are used as configuration files to link power-aware libraries with SystemC models.

Other works provide AOP features not only for verification, but also for the actual
design of hardware. \emph{D\'{e}harbe and Medeiros}~\cite{Deharbe:2006} present and assess possible
applications of AOP in the context of integrated system design by using SystemC with AspectC++.
Differently from the works discussed previously, they showed how AOP can be used to encapsulate some
functional characteristics of hardware components. They modeled as aspects the replacement policy of
a cache, the data type of an FFT, and the communication protocol between modules. However, only
simulation results are shown and they do not compare the implementation of aspect-based components
against components with all the functionalities hand-coded. In a similar work, \emph{Liu et
al.}~\cite{FengLiu:2009} implemented a SystemC model for a 128-bit floating-point adder and
described the implementation of the same model using AOP techniques. But, synthesis results are not
provided and the two models are compared only in terms of functionality to show that the AOP design
works like the original SystemC-only design. ASystemC~\cite{Endoh:2011} also extends SystemC in a
similar fashion, but, instead of using AspectC++, authors developed their own aspect weaver.
The new aspect language was introduced through different case studies involving high-level
estimation of circuit size, feature-configurable products, and assertion-based verification.
However, the evaluation of ASystemC also does not compare hardware generated using traditional
methods with their proposal.

Other works in this area follow different approaches. The \emph{e} programming
language~\cite{Vax:2007} was designed for modeling and verification of electronic systems and some
of its mechanisms can be used to support AOP features. Apart from its OOP features, \emph{e} has
some constructs to define the execution order of overloaded methods in inherited classes, which can
be used to define pointcuts and implement aspects. Indeed, this can be used to implement the
behavior of hardware components, but \emph{e} is more focused in high-level specification and there
is not any tool support for synthesis. \emph{Jun et al.}~\cite{Jun:2009} analyzed the application of
\textit{Aspectual Feature Module} ~(AFM)~\cite{Apel:2008} to HDLs. They have implemented a RISC
processor using SystemC and FeatureC++~\cite{Apel:2008}, and showed how AFM enables the incremental
development of hardware through the modularization of code fragments for the implementation of a
function. However, AOP is used only for encapsulation of verification code and the authors
do not provide synthesis results of the resulting code.

\section{Designing hardware components using scenario adapters and configurable features}
\label{PROPOSAL}

Similarly to previous works, we also based our approach on methodologies which have been used in the
software domain. The \textit{Application-driven Embedded System
Design}~(ADESD)~\cite{Froehlich:2001} methodology elaborates on commonality and variability
analysis---the well-known domain decomposition strategy behind OOP---to add the concept of aspect
identification and separation at early stages of design. Throughout ADESD's domain engineering
process, the properties that transcend the scope of single abstractions are captured as
\emph{aspects}. Such aspects include mostly dependencies from the execution scenario and
non-functional properties. This enables components to be reused on different scenarios with the
application of proper \emph{aspects}. This aspect weaving is performed by constructs called
\emph{Scenario adapters}~\cite{Froehlich:2000}. The design artifacts originally 
proposed in ADESD are already implemented and validated on the \textit{Embedded Parallel Operating
System}(EPOS)~\cite{EPOS}. EPOS aims to automate the development of dedicated computing systems,
and features a set of tools to select, adapt, and plug components into an application-specific
framework, thus enabling the automatic generation of an application-oriented system instance. EPOS
is implemented in C++ and leverages on \emph{generative programming}~\cite{Czarnecki:2000}
techniques
such as \emph{static metaprogramming} in order to achieve high reusability with low overhead.

@@

Additionally to the analysis and domain engineering process, several characteristics can be
identified by the designer as configurable features of the components. In fact, such characteristics
represent fine variations within a component, which can be set in order to change slightly its
behavior or structure.

Metaprograms modify structure through inheritance.  The example below shows how one can define the
base class of \emph{Component} as a configurable feature:

Figure \ref{fig_config_features_uml} shows how this concept can be represented in a UML diagram.
The component definition indicates its features. In the case of conditional inheritance, both
possibilities are represented with a special making at the relationship definition.

\fig{.6}{fig_config_features_uml}
{UML representation for configurable features}

@@

Scenario adapters were developed around the idea of components getting \emph{in} and \emph{out} of
an execution scenario, allowing actions to be executed at these points, therefore, a scenario must
define at least two different operations: \emph{enter} and \emph{leave}. These actions must take
place respectively before and after each of the component's operation in order to setup the
conditions required by the scenario. For example, in a compressed scenario, \emph{enter}
would be responsible to decompress the component's input data, while \emph{leave} would compress its
outputs. Figure \ref{fig_scenario_adapters_orig} shows the general structure of a scenario adapter.
The \emph{Scenario} class represents the execution scenario and incorporates, via aggregation, all
of the aspects which define its characteristics. Then, the adaptation of the component to the
scenario is performed by the \emph{Scenario Adapter} class using inheritance. In classical AOP
terms, there is a \emph{join point} before and after each method of a component's public
interface, while the \emph{pointcuts} are defined by the scenario adapter.

\fig[ht]{.6}{fig_scenario_adapters_orig}
{UML-based class diagram showing the general structure and behavior of a scenario adapter.}

@@

\fig[ht]{.6}{fig_systemc_module}
{SystemC description of a RTL component with a simple handshaking interface.}

@@

\fig[ht]{.6}{fig_sm_wait}
{A state machine implemented using \emph{SC\_CTHREAD} and \emph{SC\_METHOD}.}

By using these constructs, one can describe components susceptible for adaptation using scenario
adapters. However, other differences between software and RTL hardware design must be considered.
In software, the execution model is naturally sequential. A software OOP model may contain several
independent classes and still have a single and sequential execution flow, unless the designer
explicitly models parallelism. On the other side, in RTL hardware all modules runs in parallel,
and the operations inside the modules are also executed in parallel unless the designer explicitly
schedule them in different clock cycles. Taking this into account, we propose to define each aspect
as a single and independent hardware component. Figure \ref{fig_aspect_common} shows this
definition.

\fig[ht]{.6}{fig_aspect_common}
{Definition of aspects as a SystemC module.}

@@

\footnotesize\begin{verbatim}
If (op_req_in) {
    op_rdy_out = 0;
    static_cast<Aspect*>(this)->trigger_behavior();
    op_rdy_out = 1;
}
else {
    op_rdy_out = 1;
    static_cast<Aspect*>(this)->idle_behavior();
}
\end{verbatim}\normalsize

\emph{Aspect\_Common} defines basic input/outputs signals and a SystemC process on which the aspect
behavior is going to execute. Each class defining an aspect inherits from \emph{Aspect\_Common} and
must implement at least two methods: \emph{Aspect::trigger\_behavior} defines the behavior
executed when an operation is triggered; and \emph{Aspect::idle\_behavior} defines the
behavior executed when the aspect is in an idle state. Notice that the aspects classes derive from
template instantiations of \emph{Aspect\_Common} using themselves as template parameters. This is
known as \textit{Curiously Recurring Template Pattern}~(CRTP)~\cite{Coplien:1995}, and we use it to
avoid the use of dynamic polymorphism, which is not supported for hardware synthesis. The final form
of the scenario adapter in hardware can be seen in Figure \ref{fig_scenario_and_adapter} (for
simplicity, some details already shown in Figures \ref{fig_systemc_module}, \ref{fig_sm_wait}, and
\ref{fig_aspect_common} are omitted).

\fig{.6}{fig_scenario_and_adapter}
{UML-based diagram of the final scenario adapter. Input/output signals are omitted for simplicity.}

@@

\footnotesize\begin{verbatim}
while(op_rdy_0 &  ... & op_rdy_n)
    wait();

for all aspects {
    ...
    //set aspects inputs/outputs
    ...
    op_req_0 = 1;
    ...
    op_req_n = 1;
}
wait();
for all aspects {
    op_req_0 = 0;
    ...
    op_req_n = 0;
}
\end{verbatim}\normalsize

@@

\subsection{ADESD and classic AOP}
%Maybe merge this subsection with the discution in section
Several previous works have already discussed aspect-oriented hardware design using SystemC and
proposed solutions based on classic AOP concepts using the AspectC++ language and weaver.
Apparently,
AspectC++ provides more powerful mechanisms for aspect implementation then ADESD's scenario
adapters, especially when it comes to the definition of pointcuts. Scenario adapters were not 
designed to add behavior in specific points inside an operation. However, it is possible to
circumvent this limitation using other standard OOP and configurable features. For example, in
\emph{D\'{e}harbe and Medeiros}~\cite{Deharbe:2006} the replacement policy of a memory cache and 
the data type of an FFT are encapsulated as aspects, while a straightforward alternative would be to
use inheritance and templates parameters.

Another advantage of ADESD over the methodologies proposed previously is the implementation of its 
mechanisms, which can be realized using only standard SystemC features. No extensions to the
language or an aspect weaving tool are required. Previous works focus on tools and mechanisms that
were deployed originally for software development (e.g. AspectC++), therefore limiting its use for
the generation of synthesizable hardware. For example, AspectC++ introduces dynamic pointers and
objects that are outside the SystemC synthesizable subset\cite{systemc_subset}, thus allowing the
development of simulation-only models. This also limits the evaluation of the physical overheads
(e.g. silicon area and performance) associated to AOP.

@@

\section{Evaluation}
\label{RESULTS}
In this section we describe the experimental results for the evaluation of our approach. First, we
describe the components which are part of our case studies along with the scenario aspects that we
have identified. Then, we provide an evaluation of the designs, considering hardware synthesis
results for both area and delay.

\subsection{Case studies}
\label{CASE_STUDY}
We have analyzed a PABX application from one of our
industry partners and designed some of its basic building blocks using the proposed design
artifacts. The basic components defined as case studies are described below. All of them
were designed according to the guidelines described in section \ref{PROPOSAL}.

\textbf{Resource scheduler:} a hardware implementation of the EPOS's
scheduler~\cite{EPOS,Marcondes:2009:2}. A scheduler may perform operations both
synchronously (upon request by another component) or asynchronously (by preempting the execution of
another component). This complexity makes it a good case study. Furthermore, a hardware-implemented
scheduler reduces jitter and improves the support for real-time applications. Such characteristics
can provide a major impact in the PABX system, since each channel in the PABX central is handled in
software by a different thread.

\textbf{FIR filter:} \textit{finite impulse response}~(FIR) filters are some of the most used and
well-known blocks in digital signal processing , thus offering a high reuse potential in a wide
range of application
domains. This motivated a careful and flexible design. The main feature of our filter is the support
to both complex and real arithmetic by using configurable feature, as shown in Figure
\ref{fig_fir_uml}. The interface and MAC core are implemented separately and inherited according
to the feature set.

\fig{.6}{fig_fir_uml}
{FIR filter structure}

>>>>

%\documentclass[10pt,journal,final,letterpaper]{IEEEtran}
%\documentclass[10pt,peerreview,final,letterpaper]{IEEEtran}
%\documentclass[prodmode,acmtodaes]{acmsmall}
\documentclass{elsarticle}
%\documentclass[final]{elsarticle}
%\documentclass[preprint,review,12pt]{elsarticle}

\usepackage[latin1]{inputenc}	% for Latin languages
\usepackage[T1]{fontenc}	% for ISO and UTF characters
\usepackage[english]{babel}	% for multilingual support
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
%\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{color}
\definecolor{Red}{rgb}{.9,0,0}
\usepackage{picins}

@@

\journal{Microprocessors and Microsystems}

@@

\begin{frontmatter}

%ACM stuff
%\markboth{T. R. Mück and A. A. Fröhlich}{AOP-based RTL HW design using SystemC}

\title{Aspect-oriented RTL HW design using SystemC}

%IEEE
%\author{Tiago Rogério Mück and Antônio Augusto Fröhlich\\
%\IEEEauthorblockA{Software/Hardware Integration Lab\\
%Federal University of Santa Catarina\\
%Florianópolis, Brazil\\ 
%Email: \{tiago,guto\}@lisha.ufsc.br}
%}

%ACM
%\author{TIAGO ROGÉRIO MÜCK and ANTÔNIO AUGUSTO FRÖHLICH
%\affil{Federal University of Santa Catarina}}

%Elsevier
\author{T. R. Mück\corref{cor1}\fnref{fn1}}
\ead{tiago@lisha.ufsc.br}
\author{A. A. Fröhlich\fnref{fn1}}
\ead{guto@lisha.ufsc.br}
\address{Federal University of Santa Catarina, Florianópolis, Brazil}

\cortext[cor1]{Corresponding author}

\fntext[fn1]{Authors full address:\\
UFSC/CTC/LISHA\\
PO Box 476\\
88040-900 Florianópolis - SC - Brazil\\
Phone/Fax: +55 48 3721-9516 }

@@

%IEEE
%\begin{IEEEkeywords}
%Aspect-oriented Programming, Digital Hardware Design, Hardware Description Languages, SystemC
%\end{IEEEkeywords}

%ACM
%\category{B.6.3}{LOGIC DESIGN}{Hardware description languages}
%\terms{Design, Languages}
%\keywords{Aspect-oriented Programming, SystemC}

@@

%Elsevier
\begin{keyword}
aspect-oriented programming \sep digital hardware design \sep hardware description languages \sep SystemC
\end{keyword}

\end{frontmatter}

%ACM
%\begin{bottomstuff}
%This work is partially supported by the \emph{Coordination for Improvement of Higher Level
%Personnel}~(CAPES), under grants RH-TVD 006/2008 and 240/2008.
%
%Author's addresses: T. R. Mück and A. A. Fröhlich, Software/Hardware Integration Lab, Federal
%University of Santa Catarina, Florianópolis, Brazil, {tiago,guto}@lisha.ufsc.br.
%\end{bottomstuff}

%ACM/IEEE
%\maketitle

\section{Introduction}
\label{INTRO}
%The IEEE TII refenrences and some sentences were added tageting a submission to TII
The complexity of digital hardware design is increasing as the advances of the semiconductor
industry allow for the use of sophisticated computational resources in a wider range of
applications. This new deployment scenario is leading to a growing interest in high-level
methodologies for
digital hardware design. Solutions and methodologies that have been successfully deployed in the
scope of large-scale software systems, such as \textit{object-oriented programming}~(OOP), are being
introduced to hardware design as well. An example of a \textit{hardware description language}~(HDL)
which supports OOP is SystemC, a C++-based HDL~\cite{Panda:2001}.

HDLs are used to create descriptions of electronic circuits which can be used either for design
verification or for hardware synthesis. Differently from most software programming
languages, HDLs are intrinsically parallel and provide explicit means for describing timing.
VHDL~\cite{VHDL:2000} and Verilog~\cite{Verilog:2001} are the most widely used HDLs for 
\textit{register transfer level}~(RTL) design. In RTL, circuits are described in terms of the
operations between storage elements which are synchronized using clock signals. In this level, a
hardware design may be composed by several \emph{modules}, whose instances are interconnected to
build the system. However, in contrast to software programming languages, module communication
occurs through timed protocols and signals defined by the module input/output interface instead of
function calls.

The use of OOP-capable HDLs (e.g. SystemC) enables an increased level of flexibility and reusability
in hardware design. Object-oriented features, such as inheritance and polymorphism, allow the
stepwise refinement and generalization of hardware IPs in a way it is not possible with standard
HDLs. However, analogous to software, in hardware some system-wide cross-cutting
concerns still cannot be elegantly encapsulated. For example, in complex circuits, interconnection
of several entities is realized by introducing buses. A bus physically interacts with other
components (e.g. CPU, DMA ...), but it is difficult to use a module or an OOP class to encapsulate
the bus because its interface and arbitration method have to be implemented in every attached
component~\cite{Engel:2008}. Other examples of cross-cutting concerns in hardware design can be
found in parts of a system related to its overall functionality or the implementation of
non-functional properties (e.g. fault-tolerance and low-power mechanisms, hardware debugging through
scan chains, clock handling, ...)~\cite{Endoh:2011}. Even by using OOP in hardware, this scattered
code is hard to maintain and bugs may be easily introduced. This motivates the introduction of
\emph{aspect-oriented programming}~(AOP)~\cite{Kiczales:1997} techniques for hardware design.

AOP is an elaboration over OOP to deal with crosscutting concerns. AOP proposes the encapsulation of
these concerns in special classes called \emph{aspects}. An aspect weaving semantic is then defined
to describe when and how aspects are applied to components. The aspect weaving is defined
in terms of \emph{pointcuts} and \emph{join points}. The latter defines which points of a program
can be affected by an aspect, while the former defines when and how aspects are inserted at the
join points. Some extensions to OOP languages have been proposed to support these concepts. For
example, AspectJ~\cite{Kiczales:2001} and AspectC++~\cite{Spinczyk:2002} extend Java and C++ with
full support for AOP features. They provide both new language constructs and an aspect weaving
tool that applies aspects to the base code before it is processed by the traditional compiling
chain.

@@

In order to contribute to this scenario, in this paper we describe a digital hardware design method
which leverages on SystemC features in order to enable the implementation of hardware components
using OOP and AOP concepts. We propose the use of a design strategy that yields
components in which its dependencies from the execution scenario are encapsulated as \emph{aspects}
and \emph{configurable features}. The proposed aspect weaving semantics are implemented using only
standard C++, hence eliminating the need of extra tools and compilers.
Additionally, such features are within the SystemC synthesizable subset~\cite{systemc_subset}, thus
yielding \emph{synthesizable components}. Our method is finally illustrated by the design and
implementation of case studies extracted from an industrial \textit{Private Automatic Branch
Exchange} (PABX) application.

@@

%TODO another one for verification: \cite{Tabakov:2012}
There are also several works which proposed the use of AOP concepts mostly for 
hardware verification. \emph{Kallel et al.}~\cite{Kallel:2010} proposed the use of SystemC and
AspectC++ to implement assertion checkers. The authors focused on the
verification of
\textit{transaction-level models}~(TLM)~\cite{Cai:2003} in which transaction state updates are used
as
pointcuts. They provide a framework in which the user's verification classes extend base aspect
classes that implement the pointcuts and the verification primitives. In \emph{Vachharajani et
al.}~\cite{Vachharajani:2004} the authors developed the \textit{Liberty Structural
Specification Language}~(LSS). In LSS each module can declare instances which emit certain
events at runtime. These events behave like pointcuts of AOP. Each time a certain state is reached
or a value is computed, the instance will emit the corresponding event and user-defined aspects will
perform statistics calculation and reporting. \emph{Liu et al.}~\cite{Liu:2010} also proposed
AOP-based instrumentation, but focusing on high-level power estimation. They have developed a
methodology based on SystemC in which AspectC++ is used to define special power-aware aspects. 
These aspects are used as configuration files to link power-aware libraries with SystemC models.

Other works provide AOP features not only for verification, but also for the actual
design of hardware. \emph{D\'{e}harbe and Medeiros}~\cite{Deharbe:2006} present and assess possible
applications of AOP in the context of integrated system design by using SystemC with AspectC++.
Differently from the works discussed previously, they showed how AOP can be used to encapsulate some
functional characteristics of hardware components. They modeled as aspects the replacement policy of
a cache, the data type of an FFT, and the communication protocol between modules. However, only
simulation results are shown and they do not compare the implementation of aspect-based components
against components with all the functionalities hand-coded. In a similar work, \emph{Liu et
al.}~\cite{FengLiu:2009} implemented a SystemC model for a 128-bit floating-point adder and
described the implementation of the same model using AOP techniques. But, synthesis results are not
provided and the two models are compared only in terms of functionality to show that the AOP design
works like the original SystemC-only design. ASystemC~\cite{Endoh:2011} also extends SystemC in a
similar fashion, but, instead of using AspectC++, authors developed their own aspect weaver.
The new aspect language was introduced through different case studies involving high-level
estimation of circuit size, feature-configurable products, and assertion-based verification.
However, the evaluation of ASystemC also does not compare hardware generated using traditional
methods with their proposal. AspectVHDL~\cite{Meier:2012}, on the other hand, defines an aspect
weaving approach that can yield synthesizable code. However, using VHDL as the base language limits
the potential for reusability since OOP features are not supported.

Other works in this area follow different approaches. The \emph{e} programming
language~\cite{Vax:2007} was designed for modeling and verification of electronic systems and some
of its mechanisms can be used to support AOP features. Apart from its OOP features, \emph{e} has
some constructs to define the execution order of overloaded methods in inherited classes, which can
be used to define pointcuts and implement aspects. Indeed, this can be used to implement the
behavior of hardware components, but \emph{e} is more focused in high-level specification and there
is not any tool support for synthesis. \emph{Jun et al.}~\cite{Jun:2009} analyzed the application of
\textit{Aspectual Feature Module} ~(AFM)~\cite{Apel:2008} to HDLs. They have implemented a RISC
processor using SystemC and FeatureC++, and showed how AFM enables the incremental
development of hardware through the modularization of code fragments for the implementation of a
function. However, AOP is used only for encapsulation of verification code and the authors
do not provide synthesis results of the resulting code.

%TODO ainda ta muito corrido a explicação dos processo. Precisa justificar melhor cada uma das 
%decisões
\section{Designing hardware components using scenario adapters and configurable features}
\label{PROPOSAL}

%TODO still too "lisheiros-oriented"
Previous works that explored the use of AOP in SystemC have relied on AspectC++. We have also based
our approach on methodologies which have been used in the
software domain. The \textit{Application-driven Embedded System
Design}~(ADESD)~\cite{Froehlich:2001} methodology elaborates on commonality and variability
analysis---the well-known domain decomposition strategy behind OOP---to add the concept of aspect
identification and separation at early stages of design. Throughout ADESD's domain engineering
process, the properties that transcend the scope of single abstractions are captured as
\emph{aspects}. Such aspects include mostly dependencies from the execution scenario and
non-functional properties. This enables components to be reused on different scenarios with the
application of proper \emph{aspects}. This aspect weaving is performed by constructs called
\emph{Scenario adapters}~\cite{Froehlich:2000}.

@@

@@

Scenario adapters were developed around the idea of components getting \emph{in} and \emph{out} of
an execution scenario, allowing actions to be executed at these points, therefore, a scenario must
define at least two different operations: \emph{enter} and \emph{leave}. These actions must take
place respectively before and after each of the component's operation in order to setup the
conditions required by the scenario. For example, in a compressed scenario, \emph{enter}
would be responsible to decompress the component's input data, while \emph{leave} would compress its
outputs. Figure \ref{fig_scenario_adapters_orig} shows the general structure of a scenario adapter.
The \emph{Scenario} class represents the execution scenario and incorporates, via aggregation, all
of the aspects which define its characteristics. Then, the adaptation of the component to the
scenario is performed by the \emph{Scenario Adapter} class using inheritance. Using classical AOP
terms, there is a \emph{join point} before and after each method of a component's public
interface, while the \emph{pointcuts} are defined by the scenario adapter.

\figSC[ht]{.6}{fig_scenario_adapters_orig}
{UML-based class diagram showing the general structure and behavior of a scenario adapter.}

@@

\figSC[ht]{.6}{fig_systemc_module}
{SystemC description of a RTL component with a simple handshaking interface.}

@@

\figSC[ht]{.6}{fig_sm_wait}
{A state machine implemented using \emph{SC\_CTHREAD} and \emph{SC\_METHOD}.}

%TODO olha a primeira frase. Como essas constructs permitem isso ?
By using these constructs, one can describe components susceptible for adaptation using scenario
adapters. However, other differences between software and RTL hardware design must be considered.
In software, the execution model is naturally sequential. A software OOP model may contain several
independent classes and still have a single and sequential execution flow, unless the designer
explicitly models parallelism. On the other side, in RTL hardware all modules runs in parallel,
and the operations inside the modules are also executed in parallel unless the designer explicitly
schedules them in different clock cycles. Taking this into account, we propose to define each aspect
as a single and independent hardware component. Figure \ref{fig_aspect_common} shows this
definition.

\figSC[ht]{.6}{fig_aspect_common}
{Definition of aspects as a SystemC module.}

@@

\vbox{
\footnotesize\begin{verbatim}
If (op_req_in) {
    op_rdy_out = 0;
    static_cast<Aspect*>(this)->trigger_behavior();
    op_rdy_out = 1;
}
else {
    op_rdy_out = 1;
    static_cast<Aspect*>(this)->idle_behavior();
}
\end{verbatim}\normalsize
}

\emph{Aspect\_Common} defines basic input/outputs signals and a SystemC process on which the aspect
behavior is going to execute. Each class defining an aspect inherits from \emph{Aspect\_Common} and
must implement at least two methods: \emph{Aspect::trigger\_behavior} defines the behavior
executed when an operation is triggered; and \emph{Aspect::idle\_behavior} defines the
behavior executed when the aspect is in an idle state. Notice that the aspect classes derive from
template instantiations of \emph{Aspect\_Common} using themselves as template parameters. This is
known as \textit{Curiously Recurring Template Pattern}~(CRTP)~\cite{Coplien:1995}, and we use it to
avoid the use of dynamic polymorphism, which is not supported for hardware synthesis. The final form
of the scenario adapter in hardware can be seen in Figure \ref{fig_scenario_and_adapter} (for
simplicity, some details already shown in Figures \ref{fig_systemc_module}, \ref{fig_sm_wait}, and
\ref{fig_aspect_common} are omitted).

\figSC{.6}{fig_scenario_and_adapter}
{UML-based diagram of the final scenario adapter. Input/output signals are omitted for simplicity.}

@@

{
\vbox{
\footnotesize\begin{verbatim}
while(op_rdy_0 &  ... & op_rdy_n)
    wait();

for all aspects {
    ...
    //set aspects inputs/outputs
    ...
    op_req_0 = 1;
    ...
    op_req_n = 1;
}
wait();
for all aspects {
    op_req_0 = 0;
    ...
    op_req_n = 0;
}
\end{verbatim}\normalsize
}

@@

Additionally to the aspect separation, several characteristics can be
identified by the designer as configurable features of the components. Such characteristics
represent fine variations within a component, which can be set in order to change slightly its
behavior or structure.

Metaprogramming techniques~\cite{Czarnecki:2000} can be associated with the trait concept to also
modify the structure through inheritance. The example below shows how one can define the base class
of \emph{Component} as a configurable feature:

Figure \ref{fig_config_features_uml} shows how this concept can be represented in a UML diagram.
The component definition indicates its features. In the case of conditional inheritance, both
possibilities are represented with a special notation at the relationship definition.

\figSC{.6}{fig_config_features_uml}
{UML representation for configurable features}

%TODO I think that moving this discussion to the beginning of the section will help the reader to 
%understand that the choice for ADESD/adapters was not arbitrary 
\subsection{ADESD and classic AOP}
Several previous works have already discussed aspect-oriented hardware design using SystemC and
proposed solutions based on classic AOP concepts using AspectC++.
Apparently,
AspectC++ provides more powerful mechanisms for aspect implementation then ADESD's scenario
adapters, especially when it comes to the definition of pointcuts. Scenario adapters were not 
designed to add behavior in specific points inside an operation. However, it is possible to
circumvent this limitation using other standard OOP and the proposed configurable features. For
example, in \emph{D\'{e}harbe and Medeiros}~\cite{Deharbe:2006} the replacement policy of a memory
cache and the data type of an FFT are encapsulated as aspects, while a straightforward alternative
would be to use inheritance and templates parameters.

It is also worth mentioning that the implementation of ADESD's mechanisms can be realized using only
standard C++/SystemC features. No extensions to the
language or an aspect weaving tool are required. Previous works focused on tools and mechanisms that
were deployed originally for software development (e.g. AspectC++), therefore limiting its use for
the generation of synthesizable hardware. For example, AspectC++ may introduces dynamic pointers
and objects that are outside the SystemC synthesizable subset\cite{systemc_subset}, thus allowing
the development of simulation-only models. Although simple aspects in AspectC++ lead to simple code
transformation, previous works showed no evidence that AspectC++ woven code can be synthesized. This
limits the evaluation of the physical overheads (e.g. silicon area and performance) associated to
AOP.

@@

\section{Evaluation}
\label{RESULTS}
In this section we describe the experimental results of the evaluation of our approach. First, we
describe the components which are part of our case studies along with the scenario aspects that we
have identified. Then, we provide an evaluation of the designs, considering reusability and 
hardware synthesis results for both area and delay.

\subsection{Case studies}
\label{CASE_STUDY}
We have analyzed a PABX application from one of our
industry partners and designed some of its basic building blocks using the proposed design
artifacts. A PABX system is basically a commutation matrix that switches connections amongst
different input/output data channels. These channels are connected to phone lines (through an AD/DA
converter), tone generators, and tone detectors. The system also provides the transmission of voice
data through an Ethernet network to support \emph{voice over IP}~(VoIP) lines. The basic
components defined as case studies are described in more details below. All of them were designed
according to the guidelines described in section \ref{PROPOSAL}.

\textbf{Resource scheduler:} the PABX system is implemented over the \textit{Embedded Parallel
Operating System}(EPOS)~\cite{EPOS}. This case is a hardware implementation of the EPOS's
thread scheduler~\cite{Marcondes:2009:2}. A scheduler may perform operations both
synchronously (upon request by another component) or asynchronously (by preempting the execution of
another component). This complexity makes it a good case study. Furthermore, a hardware-implemented
scheduler reduces jitter and improves the support for real-time applications. Such characteristics
can provide a major impact in the PABX system, since each channel in the PABX central is handled in
software by a different thread.

\textbf{FIR filter:} \textit{finite impulse response}~(FIR) filters are some of the most frequently used and
well-known blocks in digital signal processing , thus offering a high reuse potential in a wide
range of application
domains. This motivated a careful and flexible design. The main feature of our filter is the support
to both complex and real arithmetic by using configurable feature, as shown in Figure
\ref{fig_fir_uml}. The interface and MAC core are implemented separately and inherited according
to the feature set.
